BayesPy: Variational Bayesian Inference in Python

### 2015-06-03 bayespy.tex
Run non-factorized Infer.NET and non-broadcasting BayesPy

The speed of the packages were compared by using two widely used models: a
Gaussian mixture model (GMM) and principal component analysis
(PCA).\footnote{The scripts for running the experiments are available as
  supplementary material.}  Both models were run for small and large
artificially generated datasets.  For GMM, the small model used 10 clusters for
200 observations with 2 dimensions, and the large model used 40 clusters for
2000 observations with 10 dimensions.  For PCA, the small model used
10-dimensional latent space for 500 observations with 20 dimensions, and the
large model used 40-dimensional latent space for 2000 observations with 100
dimensions.  Infer.NET used a fully factorizing approximation for the PCA model,
whereas BayesPy did not factorize with respect to the latent space dimensions.
The experiments were run on a Linux computer with a quad-core processor
(i7-4702MQ).  Because the packages run practically identical algorithms and thus
converged to similar solutions in a similar number of iterations (50--100
iterations depending on the dataset), we compared the average CPU time per
iteration.

The results are summarized in Table~\ref{tab:speed}.  On the smaller datasets,
Infer.NET is slightly faster because BayesPy has some Python overhead.  However,
the total CPU time for both methods is only a few seconds.  On the larger
datasets, BayesPy is significantly faster because it uses BLAS and LAPACK
libraries for fast numerical operations.\footnote{Assuming that NumPy uses
  BLAS/LAPACK because BayesPy uses NumPy for numerical operations.}  In
addition, if using multi-threaded BLAS and LAPACK on a multi-core computer, the
difference in wall-clock time can be even larger than in CPU time.  The speed of
BayesPy for these models might be improved by using collapsed or stochastic
variational inference.  The accuracy of BayesPy could be improved by using a
Gaussian-Wishart node, which does not factorize with respect to the Gaussian and
Wishart variables.

\begin{table}[tb]
  \centering
  \caption{The average CPU time (in milliseconds) per iteration for each dataset.}
  \small
  \begin{tabular}{ccccc}
    &
    Small GMM
    &
    Large GMM
    &
    Small PCA
    &
    Large PCA
    \\
    \hline
    BayesPy     & 45 & 210  & 52 & 380
    \\
    Infer.NET   & 26 & 4400 & 37 & 2200
  \end{tabular}
  \label{tab:speed}
\end{table}

>>>>

The speed of the packages were compared by using two widely used models: a
Gaussian mixture model (GMM) and principal component analysis
(PCA).\footnote{The scripts for running the experiments are available as
  supplementary material.}  Both models were run for small and large
artificially generated datasets.  For GMM, the small model used 10 clusters for
200 observations with 2 dimensions, and the large model used 40 clusters for
2000 observations with 10 dimensions.  For PCA, the small model used
10-dimensional latent space for 500 observations with 20 dimensions, and the
large model used 40-dimensional latent space for 2000 observations with 100
dimensions.
For the PCA model, Infer.NET used a fully factorizing approximation and also
similarly to BayesPy an approximation which did not factorize with respect to
the latent space dimensions.
%Infer.NET used a fully factorizing approximation for the PCA model,
%whereas BayesPy did not factorize with respect to the latent space dimensions.  
The experiments were run on a Linux computer with a quad-core processor
(i7-4702MQ).  Because the packages run practically identical algorithms and thus
converged to similar solutions in a similar number of iterations (50--100
iterations depending on the dataset), we compared the average CPU time per
iteration.

The results are summarized in Table~\ref{tab:speed}.  For GMM, BayesPy is
significantly faster, which is probably mostly explained by the fact that
BayesPy uses highly optimized numerical libraries (BLAS and LAPACK).  For PCA,
BayesPy automatically avoids redundant computations which arise because equal
posterior covariance matrices are shared between latent variables.  This
broadcasting makes BayesPy significantly faster than Infer.NET.  However,
broadcasting is not always possible (e.g., if there are missing values in the
PCA data), thus the table also presents the results when BayesPy is forced not
to utilize broadcasting.  In that case, BayesPy is a few times slower than
Infer.NET because Infer.NET uses a fully factorized approximation.  If Infer.NET
uses similar factorization as BayesPy, Infer.NET may be orders of magnitude
slower.
% On the smaller datasets,
% Infer.NET is slightly faster because BayesPy has some Python overhead.  However,
% the total CPU time for both methods is only a few seconds.  On the larger
% datasets, BayesPy is significantly faster because it uses BLAS and LAPACK
% libraries for fast numerical operations.\footnote{Assuming that NumPy uses
%   BLAS/LAPACK because BayesPy uses NumPy for numerical operations.}  In
% addition, if using multi-threaded BLAS and LAPACK on a multi-core computer, the
% difference in wall-clock time can be even larger than in CPU time.
The speed of BayesPy for both models might be improved by using collapsed or
stochastic variational inference.  The accuracy of BayesPy could be improved by
using a Gaussian-Wishart node, which does not factorize with respect to the
Gaussian and Wishart variables.

\begin{table}[tb]
  \centering
  \caption{
    The average CPU time (in milliseconds) per iteration for each
    dataset.  The results in parentheses have the following meanings: a) BayesPy without utilizing
    broadcasting. b) Infer.NET using an approximation factorized similarly as in
    BayesPy.
  }
  \small
  \begin{tabular}{ccccc}
    &
    Small GMM
    &
    Large GMM
    &
    Small PCA
    &
    Large PCA
    \\
    \hline
    BayesPy     & 5 & 85  & 7 (74) & 360 (6\,200)
    \\
    Infer.NET   & 25 & 4\,600 & 37 (350) & 2\,200 (210\,000)
  \end{tabular}
  \label{tab:speed}
\end{table}
